---
title: "Extra Bases"
output: html_document
date: "2023-11-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question:
### Identify players who are the most aggressive and successful baserunners out of the batters box.

```{r warning=FALSE}
library(mgcv)
library(tidyverse)
library(forecast)

load("G:/My Drive/Baseball/R Projects/Data/Statcast2023.RData")
```


```{r warning=FALSE}

# Create data of balls in play
hip <- c('single', 'double', 'triple')

hip_data <- Statcast2023 %>%
  filter(events %in% hip) %>%
  mutate(base = if_else(events == 'single', 1,
                           if_else(events == 'double', 2,
                                   if_else(events == 'triple', 3,
                                           NA)))) %>%
  select(player_name, events, bb_type, hc_x, hc_y, launch_angle, launch_speed, 
           base)

# Partition into training and validation data
RNGkind(sample.kind = 'Rounding')
set.seed(42)
train_rows <- sample(1:dim(hip_data)[1], dim(hip_data)[1]*0.8)
train_data <- hip_data[train_rows,]  
valid_data <- hip_data[-train_rows,]
```

### Create general additive model

GAM models non-linear relationships between related variables. This model creates a "predicted bases" metric based on the chosen variables. It is most similar to an xSLG calculation, but it only considers hits in play. In this way, it focuses on baserunning.

launch_speed: exit velocity
hc_x & hc_y: hit coordinates

```{r}
model <- gam(base ~ s(launch_speed, launch_angle, hc_x, hc_y),
             data = train_data)
summary(model)
```

```{r}
# Predict on validation data
predictions <- as.vector(predict(model, newdata = valid_data, na.action = na.pass))

pred_dat <- cbind.data.frame(valid_data$player_name, predictions, valid_data$base) %>% na.omit

accuracy <- accuracy(predictions, valid_data$base)
accuracy
```

```{r}

# Add predictions on all 2023 data
season_preds <- as.vector(predict(model, newdata = hip_data))

full_data <- cbind.data.frame(hip_data$player_name, season_preds,
                              hip_data$base) %>% na.omit %>%
  rename(name = `hip_data$player_name`,
         predicted = season_preds,
         actual = `hip_data$base`) %>%
  mutate(diff = actual - predicted)

head(full_data)
```

```{r}
# Group by player and calculate sum and mean
extra_bases <- full_data %>%
  group_by(name) %>%
  summarize(total = sum(diff),
            mean = mean(diff)) %>%
  arrange(by = desc(total))

head(extra_bases, 15) %>% arrange(by = desc(total))
tail(extra_bases, 15) %>% arrange(by = total)
```

## Conclusion

The tables above list the players who took the most and fewest "extra bases" out of the batters box in 2023. The leaders in this category can be understood as those who are the most aggressive out of the box, taking hustle doubles and pushing for triples when most do not. This is also a very reliable form of power production because it is more reflective of play style than present success. As the saying goes, "Speed never slumps." Likewise, good baserunning never slumps.

### Most Extra Bases

1. Jarren Duran (BOS)
2. TJ Friedl (CIN)
3. Corbin Carroll (ARI)
4. Bobby Witt Jr. (KC)
5. Ke'Bryan Hayes (PIT)
6. Steven Kwan (CLE)
7. Gunnar Henderson (BAL)
8. Elly De La Cruz (CIN)
9. Kevin Kiermaier (TOR)
10. Shohei Ohtani (LAA)

### Fewest Extra Bases

1. Salvador Perez (KC)
2. Luis Arraez (MIA)
3. Martín Maldonado (HOU)
4. Mitch Garver (TEX)
5. Eugenio Suárez (SEA)




